Visualising CHD Pathway Inequalities by ethnicity - DRAFT

### 2024-03-12 findings.qmd
Updated findings_draft with re-arranged order of presentation and adding title banner and logo

This report quantifies and illustrates levels of inequity across 34 metrics at various points along the continuum of coronary heart disease progression and over a typical treatment pathway. They are shown in the table below grouped by domain (risk factors, risk factor identification, primary prevention, disease identification, secondary prevention, tertiary prevention, intermediate and full outcomes), which represent the various stages along the pathway. Full definitions and data sources for each metric are included in the appendix.

@@

::: panel-tabset
## Five categories

@@

## Fourteen categories

@@

The map below presents each GP practice coloured according to their assigned cluster. Cluster 1 (red) is the least diverse and cluster 5 (orange) is the most diverse.

@@

xxxxxxxxxxxxxx

@@

The K-medoids clusterimg was performed using the {cluster} package[^7] in R using partitioning (clustering) of the data into k clusters "around medoids" (PAM), which is a more robust version of K-means.

@@

Clustering experiments were conducted using a variety of different combinations of ethnic breakdowns, such as 14 ethnic groups and 5 ethnic groups, as well as variations that excluded the % of white british patients or factored in age group. Techniques were also used to attempt to determine the optimal number of clusters such as the elbow method, gap plots, and scree plots. Consideration was given to combining some ethnic groups and in doing so a correlation matrix was produced to identify ethnic groups that commonly featured together. It was also important to seek a number of clusters that could be defined easily, where characteristics did not overlap each other so that resulting clusters were too similar to each other. It was also necessary to ensure that there was a sufficient number of practices in each cluster, too few practices in a cluster could result in issues with small amounts of metric data when calculating activity rates and consequently wide confidence intervals in the index of disparity calculations.

In conclusion, the final version selected and presented here used the percentage of the GP list size in each of the 14 census 2021 ethnic groups and five clusters was felt to give a sufficient number of describable clusters for which the index of disparity calculations could be performed. The chart below is an elbow plot generated for the 14 ethnic groups showing an elbow around 4 or 5 clusters.

@@

Table 3 shows the number of GP practices assigned to each of the 5 clusters and the characteristics of the clusters are descirbed above in the main body of the report.

@@

> "Practitioners need to be aware that statistical significance differs from practical importance in that statistical significance is highly dependent upon sample size. For a large sample, a statistically significant result is likely to be obtained even when the actual magnitude of an effect is small and of little or no practical importance. On the other hand, for a small sample, it is quite likely that insufficient evidence of a statistically significant result will be obtained – even when there is, indeed, an effect of practical importance." (Hahn, G. J., Doganaksoy, N., Meeker, W. Q., 2019)[^8]

[^8]: <https://academic.oup.com/jrssig/article/16/4/20/7038025>

@@

An explanation of bootstrap theory and confidence intervals is available in Practical Statistics for Data Scientists (Bruce, P. and Bruce, A., 2017)[^9]

[^9]: <https://www.oreilly.com/library/view/practical-statistics-for/9781491952955/>

@@

All the code used in this project is available on the Strategy Unit GitHub[^10].

[^10]: <https://github.com/The-Strategy-Unit/bhf_chd-ethnicity>

>>>>

This report quantifies and illustrates levels of inequity across 35 metrics at various points along the continuum of coronary heart disease progression and over a typical treatment pathway. They are shown in the table below grouped by domain (risk factors, risk factor identification, primary prevention, disease identification, secondary prevention, tertiary prevention, intermediate and full outcomes), which represent the various stages along the pathway. Full definitions and data sources for each metric are included in the appendix.

@@

::: panel-tabset
## Five ethnic groups

@@

## Fourteen ethnic groups

@@

The interactive map below presents each GP practice coloured according to their assigned cluster. Cluster 1 (red) is the least diverse and cluster 5 (orange) is the most diverse.

@@

The 2021 Census Ethnicity data by LSOA uses the 2021 LSOAs. The GP list size data uses the 2011 LSOAs. Some LSOAs that have gp list population assigned to them do not feature in the census data of ethnicity by lsoa. Need to check if these are the changes from 2011 to 2021 census.

Possible solution would be to work out the average proportions for all the other LSOAs for the practice and distribute the gp list population for the lsoas that have no match proportionally across the ethnic groups.

Another option would be where lsoas in 2021 have a parent lsoa in 2011 then apply the average ethnicity % for the child lsoas to the parent (this is when one 2011 lsoa has been split into multiple 2021 lsoas.) Where the lsoas in 2021 have a child lsoa in 2011 then apply the ethnicity % for the 2021 parent to the 2011 children (this is when several lsoas have been combined into one new lsoa)

join lsoa_lookup to census 21eth using the 2021 lsoa code in the lookup dataset.

Have opted for the latter solution.

Imputing data in Census 2021:

<https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/methodologies/itemeditingandimputationprocessforcensus2021englandandwales>

Variation in age groups for different ethnicities:

<https://www.ethnicity-facts-figures.service.gov.uk/uk-population-by-ethnicity/demographics/age-groups/latest/>

<https://www.ethnicity-facts-figures.service.gov.uk/uk-population-by-ethnicity/demographics/age-groups/latest/#download-the-data>

@@

The K-medoids clustering was performed using the {cluster} package[^7] in R using partitioning (clustering) of the data into k clusters "around medoids" (PAM), which is a more robust version of K-means. The algorithm is based on the search for k representative objects or medoids among the observations of the dataset. After finding a set of k medoids, k clusters are constructed by assigning each observation to the nearest medoid. The goal is to find k representative objects which minimise the sum of the dissimilarities of the observations to their closest representative object.[^8]

@@

[^8]: <https://cran.r-project.org/web/packages/cluster/cluster.pdf>

Clustering experiments were conducted using a variety of different combinations of ethnic breakdowns, such as 14 ethnic groups and 5 ethnic groups, as well as variations that excluded the % of White British patients or factored in age group. Techniques were also used to attempt to determine the optimal number of clusters such as the elbow method, gap plots, and scree plots. Consideration was given to combining some ethnic groups and in doing so a correlation matrix was produced to identify ethnic groups that commonly featured together. It was also important to seek a number of clusters that could be defined easily, where characteristics did not overlap each other so that resulting clusters were too similar to each other. It was also necessary to ensure that there was a sufficient number of practices in each cluster, too few practices in a cluster could result in issues with small amounts of metric data when calculating activity rates and consequently lead to wide confidence intervals in the index of disparity calculations.

In conclusion, the final version selected, and presented here, used the percentage of the GP list size in each of the 14 census 2021 ethnic groups and five clusters was felt to give a sufficient number of describable clusters for which the index of disparity calculations could be performed. The chart below is an elbow plot generated for the 14 ethnic groups showing an elbow at around 4 or 5 clusters.

@@

Table 3 shows the number of GP practices assigned to each of the 5 clusters and the characteristics of the clusters are described in the main body of the report above.

@@

> "Practitioners need to be aware that statistical significance differs from practical importance in that statistical significance is highly dependent upon sample size. For a large sample, a statistically significant result is likely to be obtained even when the actual magnitude of an effect is small and of little or no practical importance. On the other hand, for a small sample, it is quite likely that insufficient evidence of a statistically significant result will be obtained – even when there is, indeed, an effect of practical importance." (Hahn, G. J., Doganaksoy, N., Meeker, W. Q., 2019)[^9]

[^9]: <https://academic.oup.com/jrssig/article/16/4/20/7038025>

@@

An explanation of bootstrap theory and confidence intervals is available in Practical Statistics for Data Scientists (Bruce, P. and Bruce, A., 2017)[^10]

[^10]: <https://www.oreilly.com/library/view/practical-statistics-for/9781491952955/>

@@

All the code used in this project is available on the Strategy Unit GitHub[^11].

[^11]: <https://github.com/The-Strategy-Unit/bhf_chd-ethnicity>
